Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation

### 2024-07-14 paper/ms.tex
update from overleaf

@@

\arxiv{}{
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a]{Kevin Tan}
\author[a]{Giles Hooker}
\author[b,1]{Edward L. Ionides}
%\author[a]{Author Three}
\affil[a]{University of Pennsylvania}
\affil[b]{University of Michigan}
%\affil[c]{Affiliation Three}
% Please give the surname of the lead author for the running footer
\leadauthor{Tan}
% Please add a significance statement to explain the relevance of your work
\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements. Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method outperforms previous approaches on a challenging problem in epidemiology.}
\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: kevtan@wharton.upenn.edu}
% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Maximum Likelihood}
}

@@

\begin{abstract}
  Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood function that is discontinuous as a function of the model parameters.  We show how to embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms.  This new class permits a bias/variance tradeoff and hence a mean squared error substantially lower than the existing algorithms. We develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate. Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities. Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm show substantial improvement on current state-of-the-art methods for a challenging scientific benchmark problem.
\end{abstract}

@@

The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \cite{ionides06-pnas,ionides15}.
Likelihood-based inference is statistically efficient \cite{pawitan01} and is robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
This plug-and-play property is useful for scientific applications \cite{he10}.
Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increases.
Algorithmic advances are needed to make plug-and-play likelihood maximization for POMP models more numerically efficient.
We develop plug-and-play AD methods which provide a large computational advantage over the methods used in these previous applications.

Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, are computationally expensive \cite{corenflos21, chen24}, or require access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.

Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
However, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.

@@

MOP-$\alpha$ avoids the issue of asymptotic bias \cite{corenflos21} arising from having to drop terms arising from resampling when differentiating through the standard particle filter as in \cite{naesseth18} in a novel way.
This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
Yet, MOP-$\alpha$ can nevertheless provide strongly consistent particle and log-likelihood estimates (Theorem \ref{thm:mop-targeting}), and when $\alpha=1$, consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).

We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity (Theorem \ref{thm:mop-convergence}).
Critically, the estimator converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \cite{ionides06-pnas,ionides11} and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

@@

\section{Problem Setup}

@@

We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
The existence of this density is necessary to define the likelihood function, but plug-and-play algorithms do not have access to evaluation of this density.
Instead, they have access to a corresponding simulator, which we write as $\process_n\left(x_n\mid x_{n-1}; \theta\right)$.
We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, which we suppose can be evaluated. We write $y_n^*$ for the actual values of the observations that were observed.
We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
Superscripts $x_{n,j}^A$ of particles denote the ancestral trajectory of particle $j$ at time $n$, $a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$, and $k_j$
denotes the resample indices drawn for each particle $j$.

@@

\section{Off-Parameter Particle Filters}

@@

\subsection{Algorithm Outline}

@@

If $\alpha=1$ in Algorithm~\ref{alg:mop}, the weights, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases.
This leads to numerical instability unless $N$ is small.
For $\alpha<1$, the weights fom previous timesteps are discounted, as in Equation~(\ref{eq:weighting-scheme}).
Heuristically, $\alpha$ controls a rate of exponential decay of the memory that the filter at time $t_n$, and its resulting gradient estimate, has over the ancestral trajectories prior to $t_n$.
With $\alpha<1$, Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$.
However, we expect bias for $\theta\neq\phi$ that shrinks as $\theta$ approaches $\phi$.
This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.

@@

Our approach has similarities with \cite{svensson18}, who construct a deterministic local approximation of the likelihood at the $k$-th optimization step via saving the particles from a run at some $\theta_{k-1}$, perform a reweighting with transition and measurement density ratios to evaluate the likelihood at any sufficiently nearby $\theta$, and then use a $0$-th order optimizer such as \texttt{optim} in \texttt{R} to choose a $\theta_k$ that maximizes the likelihood in that neighborhood. 
However, due to their use of transition densities in the reweighting, their approach is not plug-and-play. The variance of their log likelihood estimate is also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is the horizon. Our approach bypasses these two issues. The use of the \textit{reparameterization trick} to consider particle paths dependent on $\theta$ (and not $\theta_{k-1}$ or $\phi$) allows us to retain the plug-and-play property. On the other hand, our use of AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$. 
The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
Our introduction of $\alpha$ avoids that difficulty.

\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}

@@

\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
    \label{thm:mop-functional-forms}
    Writing $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and using the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$, when $\alpha=0$,
    \vspace*{-2.5mm}
    \begin{equation} \nonumber
        \nabla_\theta \hat\ell^0(\theta) 
        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
        \vspace*{-2.5mm}
    \end{equation}
    yielding the estimate of \cite{naesseth18} for the bootstrap filter. When $\alpha=1$,
    \vspace*{-2.5mm}
    \begin{equation} \nonumber
        \nabla_\theta \hat{\ell}^1(\theta) 
        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
    \vspace*{-2.5mm}
    \end{equation}
    yielding the estimator of \cite{poyiadjis11, scibior21} for the bootstrap filter.
\end{thm}

We defer the proof to \arxiv{Appendix~\ref{appendix:functional}}{the supplementary material}. 
The argument relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case. 
Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.

This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
As \cite{scibior21} remarks, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step. 
This is in contrast to the case when $\alpha=1,$ as that estimate, studied by \cite{poyiadjis11}, only considers the surviving particles at time $N$ and so fully tracks dependencies over time.

\begin{figure}[ht!]
  \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/biasvar.png}
    \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}. We display the MSE of score estimates for the trend in transmission, evaluated at the MLE.}
    \label{fig:biasvar}
\end{figure}

\subsection{Summary of Theoretical Guarantees}

@@

\section{Practical Maximum-Likelihood Estimation}

@@

If $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide. One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate. This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the use of the stop-gradient operation by \cite{scibior21}.

\subsection{Optimization}

This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$. We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative (first or second-order) method that uses the MOP-$\alpha$ gradient estimate. This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima. Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood. Combining these two methods in this way lets us enjoy the best of both worlds.

@@

\subsection{Linear Convergence Rates}

On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in the theory of convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.

@@

Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $\|\nabla_\theta \hat\ell^\alpha(\theta_m)\| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$, for some $\beta \in (0,1)$. Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$. Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$. Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is an $\epsilon$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
\arxiv{}{\vspace*{-2mm}}
$$
\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big).
$$
\label{thm:mop-convergence}
\end{thm}
\arxiv{}{\vspace*{-5mm}}
We borrow from tools in the field of randomized numerical linear algebra \cite{mahoney16} to solve this problem. The proof, which is similar to that of Theorem 6 in \cite{mahoney16}, is deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$. When $\alpha=1$, the required $J$ is given in the supplementary information in Lemmas \ref{lemma:grad_bound} and \ref{lemma:hess_bound}. This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score. As , we note that it is possible.

We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold. We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE, and will explore this in future work.

@@

\begin{algorithm}[H]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
        Run IF2 until initial "convergence" under cooling schedule $\eta_m$, or for a fixed number or iterations, to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} \big(-\hat\loglik(\theta_m)\big)$, $H(\theta_m)$ s.t. $\lambda_{\min}\big(H(\theta_m)\big) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_m.$
\end{algorithm}

>>>>

\AtEndDocument{\refstepcounter{thm}\label{finalthm}}
\AtEndDocument{\refstepcounter{lem}\label{finallem}}
\AtEndDocument{\refstepcounter{defn}\label{finaldefn}}
\AtEndDocument{\refstepcounter{prop}\label{finalprop}}
\AtEndDocument{\refstepcounter{aspt}\label{finalaspt}}

@@

\arxiv{}{
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a]{Kevin Tan}
\author[a]{Giles Hooker}
\author[b,1]{Edward L. Ionides}
%\author[a]{Author Three}
\affil[a]{University of Pennsylvania}
\affil[b]{University of Michigan}
%\affil[c]{Affiliation Three}
% Please give the surname of the lead author for the running footer
\leadauthor{Tan}
% Please add a significance statement to explain the relevance of your work
\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements. Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method outperforms previous approaches on a challenging problem in epidemiology.}
\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Likelihood}
}

@@

\begin{abstract}
  Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood function that is discontinuous as a function of the model parameters.  We embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms.  This new class permits a bias/variance tradeoff and hence a mean squared error substantially lower than the existing algorithms. We develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate. Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities. Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm show substantial improvement on current state-of-the-art methods for a challenging scientific benchmark problem.
\end{abstract}

@@

The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \cite{ionides06-pnas,ionides15}.
Likelihood-based inference is statistically efficient \cite{pawitan01} and is robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
This \textit{plug-and-play property} is useful for scientific applications \cite{he10}.
Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increases.
Algorithmic advances are needed to make plug-and-play likelihood maximization for POMP models more numerically efficient.
We develop plug-and-play AD methods which provide a large computational advantage over the methods used in these previous applications.

Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, are computationally expensive \cite{corenflos21, chen24}, or require access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}. Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.

However, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.

@@

One can simply drop terms arising from resampling when differentiating through the particle filter \cite{naesseth18}, but this leads to asymptotic bias \cite{corenflos21}.
MOP-$\alpha$ is constructed as a smooth particle filter, thus bypassing the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
Yet, MOP-$\alpha$ can nevertheless provide strongly consistent particle and log-likelihood estimates (Theorem \ref{thm:mop-targeting}), and when $\alpha=1$, consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).

We derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
Critically, the estimator converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \cite{ionides06-pnas,ionides11} and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

@@

\arxiv{}{\vspace*{-2mm}}
\section{Problem Setup}

@@

We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
The existence of this density is necessary to define the likelihood function, but plug-and-play algorithms do not have access to evaluation of this density.
Instead, they have access to a corresponding simulator, which we write as $\process_n\left(x_n\mid x_{n-1}; \theta\right)$.
We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, which we suppose can be evaluated. We write $y_n^*$ for the actual values of the observations that were observed.
We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
Superscripts $x_{n,j}^A$ of particles denote the ancestral trajectory of particle $j$ at time $n$,  and $k_j$
denotes the resample indices drawn for each particle $j$. 
%$a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$,

@@

\arxiv{}{\vspace*{-2mm}}
\section{Off-Parameter Particle Filters}

@@

\arxiv{}{\vspace*{-2mm}}
\subsection{Algorithm Outline}

@@

Heuristically, $\alpha$ controls the exponentially decaying memory that the filter at time $t_n$, and its resulting gradient estimate, has over the ancestral trajectories prior to $t_n$. If $\alpha=1$ in Algorithm~\ref{alg:mop}, the gradient of the weights, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulates as $n$ increases.
This leads to numerical instability and high variance unless $N$ is small.
For $\alpha<1$, the weights from previous timesteps are discounted, as in Equation~(\ref{eq:weighting-scheme}). Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$, but has an asymptotically biased score estimate.
This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
%bias for $\theta\neq\phi$ that shrinks as $\theta$ approaches $\phi$
%

@@

Our approach has similarities with \cite{svensson18}, who reweight particles from a previous run at $\phi$ to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
However, their approach is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is exponentially increasing in the horizon. Our approach bypasses these issues -- the first with the \textit{reparameterization trick} to consider particle paths dependent only on $\theta$, and the second as using AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$. 
%The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
% Our introduction of $\alpha$ avoids that difficulty.
%also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is

\arxiv{}{\vspace*{-2mm}}
\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}

@@

\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
    \label{thm:mop-functional-forms}
    Writing $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and using the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$, when $\alpha=0$,
    \vspace*{-2.5mm}
    \begin{equation} \nonumber
        \nabla_\theta \hat\ell^0(\theta) 
        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
        \vspace*{-2.5mm}
    \end{equation}
    yielding \cite{naesseth18}'s estimator with the bootstrap filter. When $\alpha=1$,
    \vspace*{-2.5mm}
    \begin{equation} \nonumber
        \nabla_\theta \hat{\ell}^1(\theta) 
        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
    \vspace*{-2.5mm}
    \end{equation}
    yielding the estimator of \cite{poyiadjis11, scibior21} for the bootstrap filter.
\end{thm}

We defer the proof to \arxiv{Appendix~\ref{appendix:functional}}{the supplementary material}. 
The argument relies on a useful decomposition of the after-resampling estimate $L_n^{A,\theta,\alpha}$. 
Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result. 
This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
As \cite{scibior21} remarks, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step. 
In contrast, the MOP-$1$ estimate of \cite{poyiadjis11, scibior21} only considers the surviving particles at time $N$ and so fully tracks dependencies over time.

\begin{figure}[ht!]
  \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/biasvar.png}
    \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}. We display the MSE of score estimates for the trend in transmission, evaluated at the MLE.}
    \label{fig:biasvar}
    \arxiv{}{\vspace*{-1ex}}
\end{figure}

\arxiv{}{\vspace*{-2mm}}
\subsection{Summary of Theoretical Guarantees}

@@

\arxiv{}{\vspace*{-2mm}}
\section{Practical Maximum-Likelihood Estimation}

@@

If $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide. One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate. This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, coinciding with and justifying the use of the stop-gradient operation in \cite{scibior21}.

\arxiv{}{\vspace*{-2mm}}
\subsection{Optimization}

To design an effective procedure for likelihood maximization with MOP-$\alpha$, we propose a simple algorithm we call Iterated Filtering followed by Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad}. IFAD runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate. This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima. Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood. Combining these two methods in this way lets us enjoy the best of both worlds.

@@

\arxiv{}{\vspace*{-2mm}}
\subsection{Linear Convergence Rates}

On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in convex optimization theory, as we show below in Theorem \ref{thm:mop-convergence}.

@@

Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $\|\nabla_\theta \hat\ell^\alpha(\theta_m)\| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$, for some $\beta \in (0,1)$. Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$. Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$. Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is an $\epsilon$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
\arxiv{}{\vspace*{-2mm}}
$$
\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big).
$$
\label{thm:mop-convergence}
\end{thm}
\arxiv{}{\vspace*{-5mm}}
We borrow from tools in the field of randomized numerical linear algebra \cite{mahoney16} to solve this problem. The proof, which is similar to that of Theorem 6 in \cite{mahoney16}, is deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$. When $\alpha=1$, the required $J$ is given in \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score.

We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. This happens fairly often in practice, for example, when local asymptotic normality of the MLE holds. We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE, and will explore this in future work.

@@

\begin{algorithm}[H]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
        Run IF2 until initial "convergence" under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
		\hspace*{4mm} $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_m.$
\end{algorithm}

\arxiv{}{\vspace*{-2mm}}
\section{Application to a Cholera Transmission Model}
